Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-04-09 VegOzone.tex
Some more comment by Ane and rewriting of subsec ozone

\begin{table}[t]
  \caption{Past and present ozone observation sites in northern Fennoscandia. Data available on \href{http://ebas.nilu.no/}{EBAS}.}
  \label{tab:ebas_obs}
  \begin{tabular}{lllllrl}
    \tophline
    Name       & Country & ID      & \multicolumn{3}{c}{Location} & Operational\\
    &         &         & lat             & lon               & alt            &\\
    &         &         & (\unit{^\circ N}) & (\unit{^\circ E})  & (\unit{m})     &\\
    \middlehline
    Esrange    & SWE     & SE0013R & 67.83           & 21.07             & 475            & 1991 -- 2018$^1$\\
    Janiskoski & RUS     & RU0001R & 68.93           & 28.85             & 118            & 1995 -- 1997\\
    Jergul     & NOR     & NO0030R & 69.45           & 24.60             & 255            & 1997 -- 2011\\
    Karasjok   & NOR     & NO0055R & 69.467          & 25.217            & 333            & 1988 -- 1997\\
    Pallas     & FIN     & FI0096G & 67.97           & 24.12             & 565            & 1995 -- present$^2$\\
    Svanvik    & NOR     & NO0047R & 69.45           & 30.03             & 30             & 1986 -- 1996$^3$\\
    \bottomhline
  \end{tabular}
  \belowtable{$^1$ Data available until the end of 2018.\\
    $^2$ Data available until the end of 2019.\\
    $^3$ Exclusive ozone monitoring in growing seasons 2018/19 for the present study operated by NILU.} % Table Footnotes
\end{table}

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}) (1986--2019). Data taken from EBAS. The hatched areas indicate time periods with insufficient quality control \citep{NILU2003}. {\bf TODO: updata with recent Pallas data}}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

\section{Statistical analysis of environmental key variables}
\label{sec:stats}
In this section, we will evaluate atmospheric conditions in the growing seasons of 2018/19 by statistical means. We will derive climatologies for both, temperature, precipitation, global irradiance, and ozone for Svanvik (Section~\ref{subsec:climatologies}). The ozone measurements will be put into a climatological as well as regional context. We will look at anomalies of the atmospheric key variable mentioned above (Section~\ref{subsec:anomalies}).

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged temperatures (red line) are displayed with standard error of mean as error band. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc). The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the time period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

Winter and spring (November--April, except for March) are relatively dry ($\sum_m \mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_m \mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurs in the summer months, with a highest monthly averaged accumulated precipitation $\sum_m \mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a large variability throughout the years.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies peak in spring, with peak values of about $48\,\unit{ppb}$ in late April. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. The ozone climatology lags about 3 months behind the temperature climatology. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski can be explained due to the lower overall statistics.

@@

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the 2 dimensional density distributions for the thus defined northern Fennoscandia (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, daily maximum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}, and daily minimum \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} are displayed together with the $1 \sigma$ uncertainties, respectively. In addition, \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} is also shown with standard error of mean (albeit small and hardly visible). Splines have been fitted through the data for easier inter-comparison with reanalysis data.

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated start of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}.

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger ozone uptake by vegetation. \chem{CO_2} uptake coniferous trees can occur as early as day of the year (\unit{doy}) 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig06}
  \caption{Derivation of ozone climatologies. Density distributions are shown together with multi-annual mean of daily ozone concentrations \chem{[O_3]}. Splines have been fitted though daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}

>>>>

\begin{table}[t]
  \caption{Past and present ozone observation sites in northern Fennoscandia. Data available on \href{http://ebas.nilu.no/}{EBAS}.}
  \label{tab:ebas_obs}
  \begin{tabular}{lllllrl}
    \tophline
    Name       & Country & ID      & \multicolumn{3}{c}{Location} & Operational\\
    &         &         & lat             & lon               & alt            &\\
    &         &         & (\unit{^\circ N}) & (\unit{^\circ E})  & (\unit{m})     &\\
    \middlehline
    Esrange    & SWE     & SE0013R & 67.83           & 21.07             & 475            & 1991 -- 2018$^1$\\
    Janiskoski & RUS     & RU0001R & 68.93           & 28.85             & 118            & 1995 -- 1997\\
    Jergul     & NOR     & NO0030R & 69.45           & 24.60             & 255            & 1997 -- 2011\\
    Karasjok   & NOR     & NO0055R & 69.467          & 25.217            & 333            & 1988 -- 1997\\
    Pallas     & FIN     & FI0096G & 67.97           & 24.12             & 565            & 1995 -- present$^2$\\
    Svanvik    & NOR     & NO0047R & 69.45           & 30.03             & 30             & 1986 -- 1996$^3$\\
    \bottomhline
  \end{tabular}
  \belowtable{$^1$ Data available on EBAS until the end of 2018.\\
    $^2$ Data available on EBAS until the end of 2019.\\
    $^3$ Exclusive ozone monitoring in growing seasons 2018/19 for the present study operated by NILU.} % Table Footnotes
\end{table}

\begin{figure*}[t]
  %\includegraphics[width=12cm]{ozone_timeseries_fenoscandic_obs.png}
  \includegraphics[width=12cm]{fig04}
  \caption{Time series of ozone observations in northern Fennoscandia (Tab.~\ref{tab:ebas_obs}) (1986--2019). Data available from EBAS. The hatched areas indicate time periods with insufficient quality control according to \citet{NILU2003}.}
  \label{fig:ozone_timesseries_fenoscandic_obs}
\end{figure*}

@@

\section{Statistical analysis of environmental key variables}
\label{sec:stats}
In this section, we will evaluate atmospheric conditions in the growing seasons of 2018/19 by statistical means. We will derive climatologies for temperature, precipitation, global irradiance, and ozone for Svanvik (Section~\ref{subsec:climatologies}). The ozone measurements will be put into a climatological as well as regional context. We will look at anomalies of the atmospheric key variable mentioned above (Section~\ref{subsec:anomalies}).

@@

\subsubsection{Temperature, precipitation, and global irradiance}
\label{subsubsec:clim_temp_prec}
The location of Svanvik at $69.45\,\unit{^\circ N}$ suggests a subarctic climate. The derived climate diagram following Walter and Lieth shown in Fig.~\ref{fig:climatediagram} supports this very well. The diagram is based on climatological data from Svanvik (1992--2012) downloaded from LandbruksMeteorologiske Tjeneste. Monthly averaged $2\,\unit{m}$ temperatures (red line) are displayed with standard error of mean (SE) as error band. We chose SE in this case due to the large interannual variability in temperature reflected in the standard deviation. Averaged monthly accumulated precipitation (blue bars) is shown with standard deviation of mean as error bars. Months with average temperatures below freezing are denoted with a star. As can be seen, temperatures stay below freezing for 5 consecutive months, while only 2 months breach $10\,\unit{^\circ C}$ regularly (July, August), satisfying the conditions for K\"{o}ppens climate classification of a regular subarctic climate (Dfc) \citep{SD:Beck2018}. The highest monthly average temperature is $(13.1\pm 1.1)\,\unit{^\circ C}$ in July and the lowest $(-12.8\pm 2.0)\,\unit{^\circ C}$ in February. In the time period 1992--2012, the coldest measured temperature was $-45.2\,\unit{^\circ C}$ observed January 27, 1999, while the highest temperature on July 16 the same year was $29.4\,\unit{^\circ C}$.

Winter and spring (November--April, except for March) are relatively dry ($\sum_m \mathrm{Precip} < 20\,\unit{mm}$). The driest month is January with $\sum_m \mathrm{Precip} = (16.7\pm 3.0)\,\unit{mm}$. The most precipitation occurs in the summer months, with a highest monthly averaged accumulated precipitation $\sum_m \mathrm{Precip} = (58.5\pm 9.2)\,\unit{mm}$ in August. The average annual accumulated precipitation given with standard error of mean is $(383\pm 86)\,\unit{mm}$. Precipitation shows a large interannual variability.

@@

\subsubsection{Ozone}
\label{subsubsec:clim_ozone}
In the following, we will look at derived ozone climatologies from different perspectives.
The multi-annual means of daily mean ozone, for simplicity here without standard deviations, at observation sites in northern Fennoscandia (refer to Table~\ref{tab:ebas_obs} for details) are shown in Supplement Fig.~\ref{fig:ozone_climatology_fenoscandic_obs}. All climatologies except for Svanvik show peak values of about $48\,\unit{ppb}$ in late April. The spring peak values at Svanvik are lower because of the much lower altitude compared to the other monitoring sites. The annual average ozone concentration \chem{\left<[O_3]\right>} declines throughout the growing season until it reaches its minimum in late summer/fall ($24-30\,\unit{ppb}$). The turning points lie approximately in June and November. At Svanvik, \chem{\left<[O_3]\right>} is $6.6\,\unit{ppb}$ lower than on sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN). The large spread in case of Svanvik and Janiskoski is due to the lower number of observations.

@@

In Figure~\ref{fig:ozone_climatology_fenoscandic_obs_spline}, the two dimensional density distributions for the combined northern Fennoscandia data (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}a) and Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) are shown. Darker colors indicate higher probability to observe these values. On top of the density distributions, $10\,\unit{days}$ averages of daily mean \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} are displayed together with $1 \sigma$ uncertainties and SE, respectively. Splines have been fitted through the data for easier inter-comparison with reanalysis data.
From the northern Fennoscandic climatology, one can deduce several features. The build-up of the ozone spring peak takes about $7.5\,\unit{months}$ from minimum to maximum. A sharp increase starts one month ahead of the estimated begin of greening $G_\mathrm{begin}$ (see Section~\ref{subsec:do3se_parameters} for details) and is in coincidence with rising temperatures and global irradiance. We find day of the year (\unit{doy}) $\left<G_\mathrm{begin}\right> = (150 \pm 14)$ for Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} is reached in late April ($100-120\,\unit{doy}$). The depletion of ozone from maximum to minimum takes only $4.5\,\unit{months}$. The minimum occurs at $25\,\unit{ppb}$ in the beginning of September ($240-250\,\unit{doy}$).

The density distribution for Svanvik and the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier decline of the ozone peak may point to a stronger influence of vegetation on the removal of ozone. \chem{CO_2} uptake of coniferous trees can occur as early as \unit{doy} 100 \citep{TB:Kolari2007, TP:Wallin2013}. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

\begin{figure}[t]
  %\includegraphics[width=8.3cm]{ozone_climatology_fenoscandic_obs_spline}
  \includegraphics[width=8.3cm]{fig06}
  \caption{Ozone climatologies. Fennoscandia combines all available data from Jergul/Karasjok, Esrange, and Pallas. The densitiy distributions are shown together with a multi-annual mean of daily ozone concentrations \chem{[O_3]} (climatology). Splines have been fitted though daily mean \chem{[O_3]}. The climatology of daily mean ozone is shown together with $1\,\sigma$ standard deviation and standard error of the mean. Since \chem{[O_3]} are strongly correlated for sites at Jergul/Karasjok (NOR), Esrange (SWE), and Pallas (FIN) (see Fig.~\ref{fig:density_distribution}), data of these have been averaged together to derive a climatology for northern Fennoscandia. (a) Northern Fennoscandia; (b) Svanvik.}
  \label{fig:ozone_climatology_fenoscandic_obs_spline}
\end{figure}
